Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis

### 2017-05-17 article/article_v0.9.5.tex
Added corrections according to Niels Keiding's feedback

In empirical psychometric work, one often encounters data that is somehow divided into two distinct subsets. This might be due to multi-center sampling frames, or to variations in instruments, orders of questionnaire item, modes of administration etc. Often, one wishes to combine such data into one an conduct only a single analysis, but this is only meaningful if the two subsets are similar in structures, that is, if they measure the same underlying concepts. However, procedures for assessing whether such a data collapsing is allowed are usually very ad hoc or require making full model assumptions. We present a suite of three visual, non-parametric diagnostic tools for investigating differences in data structures for two datasets containing different observations of the same variables, refered to as \textit{Principal Component Analysis-based Data Structure Comparisons} (PCADSC). This approach requires no directional nor hierarchical assumptions on the data. Since the tools are all based on Principal Component Analysis (PCA), they address differences in the structures of the covariance matrices of the two datasets, and by targeting different aspects of the PCA information, the three plots give a thorough and deep understanding of where differences occur and why it might be the case.

@@

Sophisticated methods for addressing this question are available when we are willing to assume a statistical model. But this places the effort of assessing data comparability very late in the data analysis process and it makes the comparability assessment %very 
ad-hoc, as it essentially relies on the appropriateness of the modeling choices, which again depends on the structure of the data. The use of (parametric) models does thus not constitute a %valid 
general data structure comparison method, but rather a fitted-model comparison method. It addresses the interplay between the model and the data, not the data alone. However, useful tools for initial, exploratory investigations of data comparability data that help researchers identify potential problems or challenges early in the study design development are mostly absent. What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables. Simple methods like variable-by-variable tests in distributional differences suffer from the drawback that they only address marginal differences and ignore the interplay between variables. On the other hand, e.g. entry-by-entry comparisons of the two empirical correlation matrices quickly becomes unmanageable as the number of variables increase.

@@

\section{PCA-based tools for data structure comparisons}
\label{sec:pcadscintro}
As mentioned above, the purpose of PCADSC is to compare overall data structures in two subsets of a dataset. But before we can get further into describing the PCADSC tools, we must first define the exact meaning of \textit{overall structures} in this context. One such definition is the structure of the covariance matrix of the dataset. If we assume all variables in the dataset to be jointly normal with known means, the covariance matrix is a sufficient statistic for describing the simultaneous distribution of all the variables. 
%This gives it a very nice interpretation as a measure of the overall structure. If we do not accept 
And without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities that say something about the interrelations between the variables in the data. All in all, the empirical covariance matrix is a reasonable place to start looking for differences in \textit{overall data structures}.

@@

The tools presented in this paper are all based on comparing the PCA results across two different datasets that contain the same variables. We denote these two datasets by $X_1$ and $X_2$, respectively, with $X_1$ consisting of $n_1$ observations of $d$ variables $x_{11},\dotsc,x_{1 n_1} \in \RR^d$, and similarly, $X_2$ consisting of $n_2$ observations of the apparently same $d$ variables, $x_{21},\dotsc,x_{2 n_2} \in \RR^d$. %
Furthermore, let $X$ be the combined dataset consisting of all $n=n_1+n_2$ observations. For these %two
three datasets, we complete the following steps:
\begin{enumerate}
\item Within each dataset, standardize each of the variables to have mean zero and unit standard deviation. Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ and $\tilde{x}_j \in \RR^d$ for $j=1,\dotsc,n$ be the standardized datasets.
\item Form the principal component analyses
\begin{align*}
S_i &= \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$,} &
S &= \frac{1}{n-1} \sum_{j=1}^n \tilde{x}_j \tilde{x}_j^\top = \sum_{k=1}^d \lambda_k \eta_k \eta_k^\top,
\end{align*}
thereby obtaining loadings $\eta_{1k}, \eta_{2k}, \eta_k$ and variance components $\lambda_{1k}, \lambda_{2k}, \lambda_k$ for $k=1,\dotsc,d$.
\end{enumerate}
The hereby obtained PCA decompositions of the correlation matrices can then be compared. We present three diagnostics plots that are designed to shine a light on different types and levels of data structure differences. These three plots are:
\begin{description}
\item[The CE plot:] The CE (cumulative eigenvalue) plot can be used to illustrate differences in variance components, that is, in the relative importance of the directions identified by the PCA. The CE plot is accompanied by two permutation-tests, testing the hypothesis of no difference in variance components.
\item[The angle plot:] The angle plot compares both loadings and variance components at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e. loading and variance component pairs) that are most similar and most different across the two datasets
\item[The chroma plot:] The chroma plot is primarily an illustration of the loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context.
\end{description}
For the deepest understanding of the data structure differences in two datasets, we suggest using all three steps in the above order.

@@

\subsubsection{The cumulative eigenvalue plot}
The cumulative eigenvalue (CE) plot compares the variation components, i.e. the eigenvalues of the correlation matrix.
%If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset. And we would expect $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0$ for $i=1,2$ to be alike excepts sample variation.
In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
\begin{align*}
(0,0), &&
(\lambda_1,\lambda_{11}-\lambda_{12}), &&
(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), &&
\ldots, &&
\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg)
\end{align*}
This may be seen as a cumulative Bland-Altman plot \citep{AltmanBland83,LinEtAl2002} for the variation components. Note that due to the standardization, the last point will always be equal to $(d,0)$. Thus, this curve will begin and end at the x-axis. And the larger excursions it makes away from the x-axis, the less alike the cumulative variation components for the two datasets are. Moreover, a positive cumulative differences implies that dataset 1 holds more information in the first components than dataset 2 does.

@@

In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparabiltity, though \cite{Veenhoven2012} presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability, while \cite{Lolle2016} shows highly potent translation issues for the term \textit{happiness}. But the problem is beyond translation efforts: The main question really is whether or not there exist such a thing as a universal, internationally valid concept of happiness. If not, the very idea of translation is meaningless, since different aspects of psychological well-being or happiness simply do not have the same relative meaning in different cultural and socio-economic settings. This is in fact a question concerning comparability of data structures. If two countries differ e.g. in terms of how social networks are typically built and structured, with one emphasizing family relations and the other mostly focusing on other social relations, then having a weak family connection does not have the same implications in the first country as it does in the second one. More specifically, whereas in the first country, lack of familial network might be related to loneliness, lack of general social capital and isolation, in the second country, the quality of the family network might not be informative at all about other aspects of a person's social or psychological well-being. The two countries thus differ in how different aspects or measures of psychological well-being are interrelated, which is essentially a difference in data structures. And therefore, comparing the two countries in these measures is not a meaningful endeavor.

In this section, we use the PCADSC tools to unveil international differences in one aspect of happiness, namely psychological well-being. Our starting point is Denmark, a small, northern European country that has repeatedly been awarded with the title of ``happiest country in the world''  by the \textit{World Happiness Report}, most recently in 2016 \citep{WHR2016}, and we wish to investigate if this title is really meaningful at all. In order to do this, we compare the Danish ESS psychological well-being data with that of Bulgaria. Though both countries are European and thus not geographically nor culturally as far apart as some other countries might be, these two countries have previously been highlighted to be very different in terms of what defines happiness \citep{ESStopline5}. Moreover, intra-European, regional differences in the relationship between social capital and happiness have also been demonstrated \citep{Rodriguez2014}, with a much less pronounced relationship between the two in Northern European countries, compared to other European countries. In particular, interpersonal relations should play a less important role in Denmark, compared to Bulgaria. Therefore, a successful method for data comparisons should be able to detect these differences by looking at data on psychological well-being from these two countries.

We also compare the Danish data with Swedish data in order to illustrate that the PCADSC tools actually do have some discriminatory power. Denmark and Sweden are both Northern European countries and are often deemed very similar in terms of culture and history. Therefore, we expect fundamental concepts such as psychological well-being to be similar across these two countries.

@@

The ESS 2012 data contains a total of 626 variables collected from 54673 citizens of 29 countries. Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being. These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}. More details on these scales can be found in \citep{ESStopline5} and the relationship between questionnaire items and scales is summarized in Table \ref{table:items} in the appendices. We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10. For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below. This gives us $n_{DK} = 1498$ observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations. Table \ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.

@@

\begin{figure}
\centerline{\includegraphics[scale = 0.7]{essDKBGpancake234.pdf}}
\caption{A chroma plot comparing the 2nd, 3rd and 4th principal components of the Bulgiarian- and Danish psychological well-being data. The component-bars are annotated with their relative variance contributions (denoted $\tilde\sigma_i$ in the above).}
\label{plotBG.pancake}
\end{figure}

@@

The chroma blot in Figure \ref{plotBG.pancake} allows us to look closer into these components. Here, we find that the relative importance of the \textit{Community wellbeing} and \textit{Supportive relationships} scales is much larger in the Bulgarian sample than in the Danish. In the Danish data, on the other hand, we find that \textit{Vitality} and \textit{Emotional well-being} seem to play bigger roles, as they appear with larger loadings in more high-ranking components in this sample, relative to the Bulgarian.

All in all, we find that psychological well-being does \textit{not} seem to be the same concept in Bulgaria and Denmark. The two countries disagree both in how many dimensions are needed to capture the most important parts of the concept (as illustrated by the differences in eigenvalues) and in how these dimensions are then weighted among the 6 scales (as illustrated by the angle- and chroma plots). In Bulgaria, interpersonal features seem to be more informative of psychological well-being, whereas in Denmark, individual characteristic play a relatively larger role, corresponding well with the previous findings mentioned above. Thus, the datasets are fundamentally different and that we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics (\cite{ESStopline5}). Moreover, the two countries cannot be ranked in terms of which country is "the most happy", at least not by referring to psychological well-being dimensions such as those encountered here.

@@

\begin{figure}
\center
\centerline{\includegraphics[scale = 0.7]{essDKSEpancake.pdf}}
\caption{A chroma plot for comparing the loading patterns of the Danish and the Swedish subsamples. Note that for each component, the bar is annotated with its cumulative variance contribution (denoted $\tilde\sigma^c_i$ in the above), that is, how much variance can be explained by having information of this and the preceding components.}
\label{plotSE.pancake}
\end{figure}

@@

Looking at the chroma plot in Figure \ref{plotSE.pancake}, the same tale is told once again: Here, we find remarkably similar loading patterns in the first three components (which are responsible for almost 80 \% of the variance in both datasets), and slight, but increasing, differences in the remaining three components. We therefore conclude that any differences in the data structures of the Danish and the Swedish samples are related to the least important dimensions of the datasets and that these dimensions are only responsible for less than 25 \% of the variance in both datasets. In particular, this means that we can combine and compare the Danish and Swedish datasets in a meaningful way and e.g. conclude using Table \ref{tableDistr} that in general, Danes seem to be somewhat more happy than Swedes, and in particular that the least happy people in Denmark (represented by the 1st quartiles) are generally a lot happier than the least happy people in Sweden. A more thorough, statistical investigation could now be put to work on answering \textit{why} this seems to be the case.

@@

In an analysis of data from the 2012 version of the European Social Survey (ESS) project, we illustrated that the PCADSC tools help inform analyses of inter-country differences in psychological well-being and happiness. It should be noted that even though concerns about pooling data from different countries are not yet very prevalent when it comes to happiness rankings, the subject has been very well scrutinized in the field of international educational rankings. For instance, the PISA tests have repeatedly been criticized for not being meaningful objects of international comparisons, especially due to problems with differential item functions \citep{Kankaras2014,Kreiner2014,ZwitserEtAl2017} and translation problems \citep{Asil2016}. This highlights lack of data comparability as a empirically relevant and concerning issue that there is no way around: When conducting internal rankings, the issue of data structure heterogeneity should always be addressed.

Though the PCADSC toolbox comprises three diagnostic plots as of now, the PCA decomposition can of course be utilized further to bring about new PCADSC plots that illustrate more aspects of data structure differences. For instance, the scree plot (as introduced by \cite{Cattell1966}), which plots the ordered eigenvalues against their component numbers, could easily be recreated as a PCADSC tool by creating scree plots for both data subsets in the same coordinate system, yielding a simple graphical comparison. The resampling approach used in the CE plot could then be adopted, thus annotating the scree plots with resampled curves. This general approach to model-fit evaluation is very helpful in combination with permutation tests, %using cumulative residuals, 
as inspired by Lin et al. \citeyearpar{LinEtAl2002}. Similarly, the angle plot could also be annotated with p-values from a permutation test based on appropriate test statistics.

The proposed methods also have limitations. First of all, as PCA is performed after standardization of the variables, the PCADSC methods cannot detect differences in mean values nor in marginal variances - this information is thrown away as the very first step. However, such differences are of a fundamentally different nature than those we have discussed so far. Differences in mean values are often the main interest of the analysis and should of course not be regarded as a data comparability issue. Differences in marginal variances, on the other hand, can pose some modeling challenges, but none that are not generally solvable by use of random effects models. We therefore do not consider this property of PCADSC to be a major drawback.

A more prominent limitation is the fact that the PCASDC methods are only valid for variables that have a numerical interpretation. In particular, this means that the methods can not be used for data with nominal categorical variables. This limitation is inherited from the PCA method and thus, moving beyond it will entail replacing the PCA framework with a more general one. Multiple correspondence analysis (MCA) is often suggested as a categorical extension of PCA \citep{Abdi2010} and including MCA in the PCADSC tools is thus a natural next step.

>>>>

In empirical psychometric work, one often encounters data that is somehow divided into two distinct subsets. This might be due to multi-center sampling frames, or to variations in instruments, orders of questionnaire item, modes of administration etc. Often, one wishes to combine such data into one and conduct only a single analysis, but this is only meaningful if the two subsets are similar in structures, that is, if they measure the same underlying concepts. However, procedures for assessing whether such a data collapsing is allowed are usually very ad hoc or require making full model assumptions. We present a suite of three visual, non-parametric diagnostic tools for investigating differences in data structures for two datasets containing different observations of the same variables, refered to as \textit{Principal Component Analysis-based Data Structure Comparisons} (PCADSC). This approach requires no directional nor hierarchical assumptions on the data. Since the tools are all based on Principal Component Analysis (PCA), they address differences in the structures of the covariance matrices of the two datasets, and by targeting different aspects of the PCA information, the three plots give a thorough and deep understanding of where differences occur and why it might be the case.

@@

Sophisticated methods for addressing this question are available when we are willing to assume a statistical model. But this places the effort of assessing data comparability very late in the data analysis process and it makes the comparability assessment %very 
ad-hoc, as it essentially relies on the appropriateness of the modeling choices, which again depends on the structure of the data. The use of (parametric) models does thus not constitute a %valid 
general data structure comparison method, but rather a fitted-model comparison method. It addresses the interplay between the model and the data, not the data alone. However, useful tools for initial, exploratory investigations of data comparability that help researchers identify potential problems or challenges early in the study design development are mostly absent. What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables. Simple methods like variable-by-variable tests in distributional differences suffer from the drawback that they only address marginal differences and ignore the interplay between variables. On the other hand, e.g. entry-by-entry comparisons of the two empirical correlation matrices quickly become unmanageable as the number of variables increase.

@@

\section{PCA-based tools for data structure comparisons}
\label{sec:pcadscintro}
As mentioned above, the purpose of PCADSC is to compare overall data structures in two subsets of a dataset. But before we can get further into describing the PCADSC tools, we must first define the exact meaning of \textit{overall structures} in this context. One such definition is the structure of the covariance matrix of the dataset. If we assume all variables in the dataset to be jointly normal with known means, the covariance matrix is a sufficient statistic for describing the joint distribution of all the variables. 
%This gives it a very nice interpretation as a measure of the overall structure. If we do not accept 
And without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities that say something about the interrelations between the variables in the data. All in all, the empirical covariance matrix is a reasonable place to start looking for differences in \textit{overall data structures}.

@@

The tools presented in this paper are all based on comparing the PCA results across two different datasets that contain the same variables. We denote these two datasets by $X_1$ and $X_2$, respectively, with $X_1$ consisting of $n_1$ observations of $d$ variables $x_{11},\dotsc,x_{1 n_1} \in \RR^d$, and similarly, $X_2$ consisting of $n_2$ observations of the apparently same $d$ variables, $x_{21},\dotsc,x_{2 n_2} \in \RR^d$. %
Furthermore, let $X$ be the combined dataset consisting of all $n=n_1+n_2$ observations. For these %two
three datasets, we complete the following steps:
\begin{enumerate}
\item Within each dataset, standardize each of the variables to have mean zero and unit standard deviation. Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ and $\tilde{x}_j \in \RR^d$ for $j=1,\dotsc,n$ be the standardized datasets.
\item Form the principal component analyses
\begin{align*}
S_i &= \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$,} &
S &= \frac{1}{n-1} \sum_{j=1}^n \tilde{x}_j \tilde{x}_j^\top = \sum_{k=1}^d \lambda_k \eta_k \eta_k^\top,
\end{align*}
thereby obtaining loadings $\eta_{1k}, \eta_{2k}, \eta_k$ and variance components $\lambda_{1k}, \lambda_{2k}, \lambda_k$ for $k=1,\dotsc,d$.
\end{enumerate}
The hereby obtained PCA decompositions of the correlation matrices can then be compared. We present three diagnostic plots that are designed to shine a light on different types and levels of data structure differences. These three plots are:
\begin{description}
\item[The CE plot:] The CE (cumulative eigenvalue) plot can be used to illustrate differences in variance components, that is, in the relative importance of the directions identified by the PCA. The CE plot is accompanied by two permutation-tests, testing the hypothesis of no difference in variance components.
\item[The angle plot:] The angle plot compares both loadings and variance components at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e. loading and variance component pairs) that are most similar and most different across the two datasets
\item[The chroma plot:] The chroma plot is primarily an illustration of the loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context.
\end{description}
For the deepest understanding of the data structure differences in two datasets, we suggest using all three steps in the above order.

@@

\subsubsection{The cumulative eigenvalue plot}
The cumulative eigenvalue (CE) plot compares the variation components, i.e. the eigenvalues of the correlation matrix. These eigenvalues represent how much information is withheld in each component in terms of explained variance. Thus, by comparing cumulative sums of variance components, it is possible to obtain a detailed picture of how the two datasets differ in terms of what components are the most informative. 
%If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset. And we would expect $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0$ for $i=1,2$ to be alike excepts sample variation.
In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
\begin{align*}
(0,0), &&
(\lambda_1,\lambda_{11}-\lambda_{21}), &&
(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), &&
\ldots, &&
\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg)
\end{align*}
%This may be seen as a cumulative Bland-Altman plot \citep{AltmanBland83,LinEtAl2002} for the variation components.
%Anne: NK pointed out that this is somewhat confusing for people who are not familiar with the Bland-Altman plot, and as far as I can tell, what we do is quite different anyway, so I think no reference is really needed(?) 
Note that due to the standardization, the last point will always be equal to $(d,0)$. Thus, this curve will begin and end at the x-axis. And the larger excursions it makes away from the x-axis, the less alike the cumulative variation components for the two datasets are. Moreover, a positive cumulative difference implies that dataset 1 holds more information in the first components than dataset 2 does.

@@

In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparabiltity, though \cite{Veenhoven2012} presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability, while \cite{Lolle2016} show highly potent translation issues for the term \textit{happiness}. But the problem is beyond translation efforts: The main question really is whether or not there exist such a thing as a universal, internationally valid concept of happiness. If not, the very idea of translation is meaningless, since different aspects of psychological well-being or happiness simply do not have the same relative meaning in different cultural and socio-economic settings. This is in fact a question concerning comparability of data structures. If two countries differ e.g. in terms of how social networks are typically built and structured, with one emphasizing family relations and the other mostly focusing on other social relations, then having a weak family connection does not have the same implications in the first country as it does in the second one. More specifically, whereas in the first country, lack of familial network might be related to loneliness, lack of general social capital and isolation, in the second country, the quality of the family network might not be informative at all about other aspects of a person's social or psychological well-being. The two countries thus differ in how different aspects or measures of psychological well-being are interrelated, which is essentially a difference in data structures. And therefore, comparing the two countries in these measures is not a meaningful endeavor.

In this section, we use the PCADSC tools to examine international differences in one aspect of happiness, namely psychological well-being. Our starting point is Denmark, a small, Northern European country that has repeatedly been awarded with the title of ``happiest country in the world''  by the \textit{World Happiness Report}, most recently in 2016 \citep{WHR2016}, and we wish to investigate if this title is really meaningful at all. In order to do this, we compare the Danish ESS psychological well-being data with that of Bulgaria. Though both countries are European and thus neither very far apart geographically nor culturally, these two countries have previously been highlighted to be very different in terms of what defines happiness \citep{ESStopline5}. Moreover, intra-European, regional differences in the relationship between social capital and happiness have also been demonstrated \citep{Rodriguez2014}. When comparing Northern European countries to other European countries, a much less pronounced relationship between the two concepts is found. In particular, interpersonal relations play a less important role in Denmark, compared to Bulgaria. Therefore, a successful method for data comparisons should be able to detect these differences by looking at data on psychological well-being from these two countries.

We also compare the Danish data with Swedish data in order to investigate if the PCADSC tools actually do hold discriminatory power. Denmark and Sweden are both Northern European countries and are often deemed very similar in terms of culture and history. Therefore, we expect fundamental concepts, such as psychological well-being, to be similar across these two countries.

@@

The ESS 2012 data contain a total of 626 variables collected from 54673 citizens of 29 countries. Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being. These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}. More details on these scales can be found in \citep{ESStopline5} and the relationship between questionnaire items and scales is summarized in Table \ref{table:items} in the appendices. We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10. For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below. This corresponds to approximately 9\% of the observations in the Danish sample, 20\% in the Bulgarian sample and only 6\% in the Swedish sample. All in all, we have $n_{DK} = 1498$ complete case observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations. Table \ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.

@@

\begin{figure}
\centerline{\includegraphics[scale = 0.7]{essDKBGpancake234.pdf}}
\caption{A chroma plot comparing the 2nd, 3rd and 4th principal components of the Bulgiarian- and Danish psychological well-being data. The component-bars are annotated with their relative variance contributions (denoted $\tilde\sigma$ in the previous section).}
\label{plotBG.pancake}
\end{figure}

@@

The chroma plot in Figure \ref{plotBG.pancake} allows us to look closer into these components. Here, we find that the relative importance of the \textit{Community wellbeing} and \textit{Supportive relationships} scales is much larger in the Bulgarian sample than in the Danish. In the Danish data, on the other hand, we find that \textit{Vitality} and \textit{Emotional well-being} seem to play bigger roles, as they appear with larger loadings in more high-ranking components in this sample, relative to the Bulgarian.

All in all, we find that psychological well-being does \textit{not} seem to be the same concept in Bulgaria and Denmark. The two countries disagree both in how many dimensions are needed to capture the most important parts of the concept (as illustrated by the differences in eigenvalues) and in how these dimensions are then weighted among the 6 scales (as illustrated by the angle- and chroma plots). In Bulgaria, interpersonal features seem to be more informative of psychological well-being, whereas in Denmark, individual characteristic play a relatively larger role, corresponding well with the previous findings mentioned above. Thus, the datasets are fundamentally different and we should therefore be wary about combining them in a joint analysis, which was also the conclusion of the ESS authors, though based on country-level aggregated statistics (\cite{ESStopline5}). Moreover, this also implies that two countries cannot be ranked in terms of which country is "the most happy", at least not by referring to psychological well-being dimensions such as those encountered here.

@@

\begin{figure}
\center
\centerline{\includegraphics[scale = 0.7]{essDKSEpancake.pdf}}
\caption{A chroma plot for comparing the loading patterns of the Danish and the Swedish subsamples. Note that for each component, the bar is annotated with its cumulative variance contribution (denoted $\tilde\sigma^c$ in the previous section), that is, how much variance can be explained by having information of this and the preceding components.}
\label{plotSE.pancake}
\end{figure}

@@

Looking at the chroma plot in Figure \ref{plotSE.pancake}, the same tale is told once again: Here, we find remarkably similar loading patterns in the first three components (which are responsible for almost 80\% of the variance in both datasets), and slight, but increasing, differences in the remaining three components. We therefore conclude that any differences in the data structures of the Danish and the Swedish samples are related to the least important dimensions of the datasets and that these dimensions are only responsible for less than 25\% of the variance in both datasets. In particular, this means that we can combine and compare the Danish and Swedish datasets in a meaningful way and e.g. conclude using Table \ref{tableDistr} that in general, Danes seem to be somewhat more happy than Swedes, and in particular that the least happy people in Denmark (represented by the 1st quartiles) are generally a lot happier than the least happy people in Sweden. A more thorough, statistical investigation could now be put to work on answering \textit{why} this seems to be the case.

@@

In an analysis of data from the 2012 version of the European Social Survey (ESS) project, we illustrated that the PCADSC tools help inform analyses of inter-country differences in psychological well-being and happiness. It should be noted that even though concerns about pooling data from different countries are not yet very prevalent when it comes to happiness rankings, the subject has been very well scrutinized in the field of international educational rankings. For instance, the PISA tests have repeatedly been criticized for not being meaningful objects of international comparisons, especially due to problems with differential item functioning \citep{Kankaras2014,Kreiner2014,ZwitserEtAl2017} and translation problems \citep{Asil2016}. This highlights lack of data comparability as an empirically relevant and concerning issue that there is no way around: When conducting internal rankings, the issue of data structure heterogeneity should always be addressed.

Though the PCADSC toolbox comprises three diagnostic plots as of now, the PCA decomposition can of course be utilized further to bring about new PCADSC plots that illustrate more aspects of data structure differences. For instance, the scree plot (as introduced by \cite{Cattell1966}), which plots the ordered eigenvalues against their component numbers, could easily be recreated as a PCADSC tool by creating scree plots for both data subsets in the same coordinate system, yielding a simple graphical comparison. The resampling approach used in the CE plot could then be adopted, thus annotating the scree plots with resampled curves. This general approach to model-fit evaluation is very helpful in combination with permutation tests, %using cumulative residuals, 
as inspired by Lin et al. \citeyearpar{LinEtAl2002}. Similarly, the angle plot could also be annotated with $p$-values from a permutation test based on appropriate test statistics.

The proposed methods also have limitations. First of all, as PCA is performed after standardization of the variables, the PCADSC methods cannot detect differences in neither mean values nor in marginal variances - this information is thrown away as the very first step. However, such differences are of a fundamentally different nature than those we have discussed so far. Differences in mean values are often the main interest of the analysis and should of course not be regarded as a data comparability issue. Differences in marginal variances, on the other hand, can pose some modeling challenges, but none that are not generally solvable by use of random effects models. We therefore do not consider this property of PCADSC to be a major drawback.

A more prominent limitation is the fact that the PCASDC methods are only valid for variables that have a numerical interpretation. In particular, this means that the methods cannot be used for data with nominal categorical variables. This limitation is inherited from the PCA method and thus, moving beyond it will entail replacing the PCA framework with a more general one. Multiple correspondence analysis (MCA) is often suggested as a categorical extension of PCA \citep{Abdi2010} and including MCA in the PCADSC tools is thus a natural next step.
